Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-10-19 text/supp.tex
Clean up supp.tex
Edits to SOM

Our prior on $\nu$ was based on \citet{fernandez1998}; they chose a more
informative exponential rate parameter of $0.1$. We chose a less informative
rate parameter of $0.01$ and truncated the distribution at $2$, since at $\nu
< 2$ the variance of the t distribution is undefined. This prior gives only
a $7.7$\% probability $\nu < 10$ but constrains the sampling sufficiently to
avoid wandering off towards infinity --- above approximately $\nu = 20$ the
t distribution is so similar to the normal distribution
(Fig.~\ref{fig:didactic}) that time series of the length considered here are
unlikely to be sufficiently informative about the precise value of $\nu$. In the
scenario where the data are not heavy tailed (e.g.~Fig.~\ref{fig:didactic}e, h)
the posterior will approximately match the prior (median $= 71$, mean $= 102$)
and not be flagged as likely heavy tailed using the metrics we used in our paper
(e.g.~Pr$(\nu < 10) > 0.5$).

@@

We separated our simulation into these two components to avoid confounding two
issues. (1) With smaller sample sizes, there may not be a stochastic draw from
the tails of a distribution. In that case, no model, no matter how perfect
a model, will be able to detect the shape of the tails. (2) Models (particularly
complex models) may return biased parameter estimates if there are conceptual,
computational, or coding errors in the models. Our first simulation tested the
first issue and our second simulation tested the latter. In general, out
simulations show that, if anything, our model under predicts the magnitude and
frequency of heavy tailed events --- especially given the length of the time
series in the GPDD.

@@

We found that we could consistently and precisely recover median posterior
estimates of $\nu$ near the true value of $\nu$ with large samples ($\ge 200$)
(Fig.~\ref{fig:sim-gompertz} upper panels). At smaller samples the we could
still usually distinguish ``heavy'' from ``not-heavy'' tails, but the model
tended to underestimate how heavy the tails were. At the same time, when the
model underestimated how heavy-tailed the distribution was, it tended to
overestimate how large the scale parameter was (Fig.~\ref{fig:sim-gompertz}
lower panels).

@@

In the second part of our simulation testing, we tested the ability of the
heavy-tailed Gompertz model to obtain unbiased parameter estimates given that
process deviations were provided in which our simpler model (in the previous
section) could accurately estimate the value of $\nu$. To generate these process
errors, we repeatedly generated candidate process deviations and estimated the
central tendency, scale, and $\nu$ values each time. We recorded when
$\hat{\nu}$ was within $0.2$ CVs (coefficient of variations) of the true $\nu$
value and used this set of random seed values in our Gompertz simulation. We
then fitted our Gompertz models to the simulated datasets with all parameters
(except $\nu$) set near the median values estimated in the GPDD. We repeated
this with $50$ and $100$ samples without observation error, $50$ samples with
observation error ($\sigma_\mathrm{obs} = 0.2$), and $50$ samples with the same
observation error and a Gompertz model that allowed for correctly specified
observation error magnitude.

@@

Observation error can bias parameter estimates \citep[e.g.][]{knape2012} and is
known to affect the ability to detect extreme events \citep{ward2007}. However,
simultaneously estimating observation and process error in state-space is
a challenging problem and is known to result in identifiability issues with the
Gompertz population model \citep{knape2008}. Furthermore, our model estimates an
additional parameter --- the shape of the process error distribution tails ---
potentially making identifiability and computational issues even greater.
Therefore, we considered a version of the base Gompertz model that allowed for
a fixed level of observation error:
\begin{align*}
U_t &= \lambda + b U_{t-1} + \epsilon_t\\
x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs})\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma_\mathrm{proc}),
\end{align*}
where $U$ represents the unobserved state vector, and $\sigma_\mathrm{obs}$
represents the standard deviation of observation error (on a log scale). We set
$\sigma_\mathrm{obs}$ to $0.2$, which represents the upper end of values often
used in simulation analyses \citep[e.g.][]{valpine2002, thorson2014b}.

@@

%\begin{figure}[htbp]
%\begin{center}
%\includegraphics[width=\textwidth]{../analysis/log10-assumed.pdf}
%\caption{Populations in which the abundance was specified as or assumed to be
%recorded as log10 values. The numbers at the front of each panel label are the
%GPDD IDs.}
    %\label{fig:log10-assumed}
%\end{center}
%\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-prior-exp0point01.pdf}
\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
\caption{
  Testing the ability to estimate $\nu$ (top panels) and the scale parameter of
  the process error (bottom panels) for a given number of samples (columns)
  drawn from a distribution with a given true $\nu$ value (rows). The red lines
  indicate the true population value. When a small number of samples are drawn
  there may not be samples sufficiently far into the tails to recapture the
  true $\nu$ value; however, heavy tails are still distinguished from normal
  tails in most cases, even with only 25 or 50 samples. TODO switch to nu
  = black swan, 3, 5, and normal.
}
\label{fig:sim-nu}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/sim-gompertz.pdf}
\caption{
  Simulation testing the Gompertz estimation model when the process error draws
  are chosen so that $\nu$ can be estimated close to the true value outside the
  full population model (``effective $\nu$'' within a CV of 0.2 of specified
  $\nu$). The simulation was run across population $\nu$ values (columns) and
  different scenarios (colours): (1) 100 time steps and no observation error,
  (2) 50 time steps and no observation error, (3) 50 time steps and observation
  error drawn from $\mathrm{Normal} (0, 0.2^2)$ but ignored, and (4) 50 time
  steps with observation error in which the quantity of observation error was
  assumed known. Dashed horizontal lines show the true population values; these
  true values were chosen to represent approximately the median values as
  estimated from the GPDD. Individual dots and lines represent a stochastic
  draws from the true population distribution and a model fitting (median and
  80\% credible intervals).
  %The panels show true values of $\nu$ =  3, 5, normal (very heavy, heavy, and
  %not heavy tailed).
  The panels from top to bottom show $1/\nu$, process noise scale parameter
  $\sigma$, growth rate parameter $\lambda$, and the density dependence
  parameter $b$.
}
\label{fig:sim-gompertz}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
%\includegraphics[width=0.7\textwidth]{../analysis/admb-coefs}
%\includegraphics[width=0.8\textwidth]{../analysis/length-vs-prob-tails}
\includegraphics[width=0.5\textwidth]{../analysis/stan-beta-correlates}
\caption{
  %Potential correlates of heavy-tailed dynamics.
  %(Top panels) Dots, thick lines, and thin lines represent fixed-effect estimates and their 50\% and 95\% confidence intervals for the covariates listed on the left: $\log \sigma$, $b$, $\lambda$, $\log$(lifespan), $\log$(data set length).
  Main effect posterior densities for potential correlates of heavy-tailed
  dynamics. The response variable was Pr($\nu < 10$) (approximately the
  probability of heavy tails). The beta regression model was fit on a logit
  scale with random intercepts for taxonomic class, taxonomic order, and
  species. All covariates were standardized by subtracting their mean and
  dividing by twice their standard deviation. Therefore, the coefficient values
  represent the expected effect of a change in two standard deviations of the
  raw input variable on the logit-transformed response variable.
}
    \label{fig:correlate-coefs}
\end{center}
\end{figure}

@@

% \clearpage
%
% \begin{figure}[htbp]
% \begin{center}
% \includegraphics[width=0.8\textwidth]{../analysis/length-vs-prob-tails}
% \caption{Model predictions. TODO}
%     \label{fig:length-predictions}
% \end{center}
% \end{figure}

>>>>

Our prior on $\nu$ was based on \citet{fernandez1998}; they chose a more
informative exponential rate parameter of $0.1$. We chose a less informative
rate parameter of $0.01$ and truncated the distribution at $2$, since at $\nu
< 2$ the variance of the t distribution is undefined. This prior gives only
a $7.7$\% probability $\nu < 10$ but constrains the sampling sufficiently to
avoid wandering off towards infinity --- above approximately $\nu = 20$ the
t distribution is so similar to the normal distribution
(Fig.~\ref{fig:didactic}) that time series of the length considered here are
unlikely to be informative about the precise value of $\nu$. In the
scenario where the data are not heavy tailed (e.g.~Fig.~\ref{fig:didactic}e, h)
the posterior will approximately match the prior (median $= 71$, mean $= 102$)
and not be flagged as likely heavy tailed using the metrics we used in our paper
(e.g.~Pr$(\nu < 10) > 0.5$).

@@

We separated our simulation into these two components to avoid confounding two
issues. (1) With smaller sample sizes, there may not be a stochastic draw from
the tails of a distribution. In that case, no model, no matter how perfect,
will be able to detect the shape of the tails. (2) Complex models may return
biased parameter estimates if there are conceptual, computational, or coding
errors. Our first simulation tested the first issue and our second simulation
tested the latter. In general, out simulations show that, if anything, our
model under predicts the magnitude and frequency of heavy tailed events ---
especially given the length of the time series in the GPDD.

@@

We found that we could consistently and precisely recover median posterior
estimates of $\nu$ near the true value of $\nu$ with large samples ($\ge 200$)
(Fig.~\ref{fig:sim-gompertz} upper panels). At smaller samples we could still
usually distinguish ``heavy'' from ``not-heavy'' tails, but the model tended to
underestimate how heavy the tails were. At the same time, the model tended to
overestimate how large the scale parameter was (Fig.~\ref{fig:sim-gompertz}
lower panels).

@@

In the second part of our simulation testing, we tested the ability of the
heavy-tailed Gompertz model to obtain unbiased parameter estimates given that
process deviations were provided in which our simpler model (previous
section) could accurately estimate the value of $\nu$. To generate these process
deviations, we repeatedly generated candidate process deviations and estimated the
central tendency, scale, and $\nu$ values each time. We recorded when
$\hat{\nu}$ was within $0.2$ CVs (coefficient of variations) of the true $\nu$
value and used this set of random seed values in our Gompertz simulation. We
then fitted our Gompertz models to the simulated datasets with all parameters
(except $\nu$) set near the median values estimated in the GPDD. We repeated
this with $50$ and $100$ samples without observation error, $50$ samples with
observation error ($\sigma_\mathrm{obs} = 0.2$), and $50$ samples with the same
observation error and a Gompertz model that allowed for correctly specified
observation error magnitude.

@@

Observation error can bias parameter estimates \citep[e.g.][]{knape2012} and is
known to affect the ability to detect extreme events \citep{ward2007}. However,
simultaneously estimating observation and process noise is a challenging
problem and is known to result in identifiability issues with the Gompertz
population model \citep{knape2008}. Furthermore, our model estimates an
additional parameter --- the shape of the process deviation tails ---
potentially making identifiability and computational issues even greater.
Therefore, we considered a version of the base Gompertz model that allowed for
a fixed level of observation error:
\begin{align*}
U_t &= \lambda + b U_{t-1} + \epsilon_t\\
x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs})\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma_\mathrm{proc}),
\end{align*}
where $U$ represents the unobserved state vector, and $\sigma_\mathrm{obs}$
represents the standard deviation of observation error (on a log scale). We set
$\sigma_\mathrm{obs}$ to $0.2$, which represents the upper end of values often
used in simulation analyses \citep[e.g.][]{valpine2002, thorson2014b}.

@@

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-prior-exp0point01.pdf}
\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
\caption{
  Testing the ability to estimate $\nu$ (top panels) and the scale parameter of
  the process deviations (bottom panels) for a given number of samples (columns)
  drawn from a distribution with a given true $\nu$ value (rows). The red lines
  indicate the true population value. When a small number of samples are drawn
  there may not be samples sufficiently far into the tails to recapture the
  true $\nu$ value; however, heavy tails are still distinguished from normal
  tails in most cases, even with only 25 or 50 samples. TODO switch to nu
  = black swan, 3, 5, and normal.
}
\label{fig:sim-nu}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/sim-gompertz.pdf}
\caption{
  Simulation testing the Gompertz estimation model when the process deviation
  draws were chosen so that $\nu$ can be estimated close to the true value
  outside the full population model (``effective $\nu$'' within a CV of 0.2 of
  specified $\nu$). The simulation was run across population $\nu$ values
  (columns) and different scenarios (colours): (1) 100 time steps and no
  observation error, (2) 50 time steps and no observation error, (3) 50 time
  steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but
  ignored, and (4) 50 time steps with observation error in which the quantity
  of observation error was assumed known. Dashed horizontal lines show the true
  population values; these true values were chosen to represent approximately
  the median values as estimated from the GPDD. Individual dots and lines
  represent a stochastic draws from the true population distribution and
  a model fitting (median and 80\% credible intervals). The panels from top to
  bottom show $1/\nu$, process noise scale parameter $\sigma$, growth rate
  parameter $\lambda$, and the density dependence parameter $b$.
}
\label{fig:sim-gompertz}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.5\textwidth]{../analysis/stan-beta-correlates}
\caption{
  Main effect posterior densities for potential correlates of heavy-tailed
  dynamics. The response variable was Pr($\nu < 10$) (approximately the
  probability of heavy tails). The beta regression model was fit on a logit
  scale with random intercepts for taxonomic class, taxonomic order, and
  species. All covariates were standardized by subtracting their mean and
  dividing by twice their standard deviation. Therefore, the coefficient values
  represent the expected effect of a change in two standard deviations of the
  raw input variable on the logit-transformed response variable.
}
    \label{fig:correlate-coefs}
\end{center}
\end{figure}

@@
